---
title: "HDAT9600 Final Assignment"
subtitle: "Submission deadline is 5pm Friday 23 2019 "
author: "insert your team name here"
date: "insert date of completion here"
output: html_document
---

```{r setup, include=FALSE}
# leave this code here, but feel free to adjust the options or add some more
# see the knitr documentation for details
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=12)
```

## Instructions

This file (hdat9600_final_assignment.Rmd) is the R Markdown document in which you need to complete your HDAT9600 final assignment. This assignment is assessed and will count for 30% of the total course marks. The assignment comprises two tasks worth 15 marks each. The first task will focus on logistic regression, and the second task will focus on survival analysis. There is no word limit, but a report of about 10 pages in length when printed (except that it will not be printed) is appropriate.

Don't hesitate to ask the course convenor for help via OpenLearning. The course instructor are happy to point you in the right direction and to make suggestions, but they won't, of course, complete your assignments for you!


## Data for this assignment

The data used for this assignment consist of records from Intensive Care Unit (ICU) hospital stays in the USA. All patients were adults who were admitted for a wide variety of reasons. ICU stays of less than 48 hours have been excluded. 

The source data for the assignment are data made freely available for the 2012 MIT PhysioNet/Computing for Cardiology Challenge. Details are provided [here](https://physionet.org/challenge/2012/). Training Set A data have been used. THe original data has been modified and assembled to suit the purpose of this assignment. Full details of the preparatory work can be found in the hdat9600-final-assignment-data-preparation file.

The dataframe consists of 120 variables, which are defined as follows:

####Patient Descriptor Variables
<li> <em>RecordID:</em>   a unique integer for each ICU stay</li>
<li> <em>Age:</em>                 years</li>
<li> <em>Gender:</em>              male/female</li>
<li> <em>Height:</em>              cm</li>
<li> <em>ICUType:</em>             Coronary Care Unit; Cardiac Surgery Recovery Unit; Medical ICU; Surgical ICU</li>
<li> <em>Length_of_stay:</em>      The number of days between the patient’s admission to the ICU and the end of hospitalisation</li>
<li> <em>Survival:</em>            The number of days between ICU admission and death for patients who died</li>


####Outcome Variables
<li> <em>in_hospital_death:</em>   0:survivor/1:died in-hospital **this is the outcome variable for Task 1: Logistic Regression**</li>
<li> <em>Status:</em>              True/False **this is the censoring variable for Task 2: Survival Analysis**</li>
<li> <em>Days:</em>                Length of survival (in days) **this is the survival time variable for Task 2: Survival Analysis**</li>


####Clinical Variables
<em>Use the hyperlinks below to find out more about the clinical meaning of each variable.</em>
The first two clinical variables are summary scores that are used to assess patient condition and risk.

<li> <em>SAPS-I score</em> [Simplified Acute Physiological Score (<a href="http://www.ncbi.nlm.nih.gov/pubmed/6499483"
target="_blank"><em>Le Gall et al., 1984</em></a>)]</li>
<li> <em>SOFA score</em> [Sequential Organ Failure Assessment (<a href="http://www.ncbi.nlm.nih.gov/pubmed/11594901"
target="_blank"><em>Ferreira et al., 2001</em></a>)]</li>

###

The following 36 clinical measures were assessed at multiple timepoints during each patient's ICU stay. For each of the 36 clinical measures, you are given 3 summary variables: a) The minimum value during the first 24 hours in ICU (_min), b) The maximum value during the first 24 hours in ICU (_max), and c) The difference between the mean and the most extreme values during the first 24 hours in ICU (_diff). For example, for the clinical measure Cholesterol, these three variables are labelled 'Cholesterol_min', 'Cholesterol_max', and 'Cholesterol_diff'.

<li><a href="http://en.wikipedia.org/wiki/Human_serum_albumin" target="_blank">
<em>Albumin</em></a> (g/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Alkaline_phosphatase" target="_blank">
<em>ALP</em></a> [Alkaline phosphatase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Alanine_transaminase" target="_blank">
<em>ALT</em></a> [Alanine transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Aspartate_transaminase" target="_blank">
<em>AST</em></a> [Aspartate transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bilirubin" target="_blank">
<em>Bilirubin</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/BUN" target="_blank">
<em>BUN</em></a> [Blood urea nitrogen (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Cholesterol" target="_blank">
<em>Cholesterol</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_creatinine#Plasma_creatinine"
target="_blank">
<em>Creatinine</em></a> [Serum creatinine (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>DiasABP</em></a> [Invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/FIO2" target="_blank">
<em>FiO2</em></a> [Fractional inspired O<sub>2</sub> (0-1)]</li>
<li><a href="http://en.wikipedia.org/wiki/Glasgow_coma_score" target="_blank">
<em>GCS</em></a> [Glasgow Coma Score (3-15)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_glucose" target="_blank">
<em>Glucose</em></a> [Serum glucose (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bicarbonate#Diagnostics"
target="_blank">
<em>HCO3</em></a> [Serum bicarbonate (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hematocrit" target="_blank">
<em>HCT</em></a> [Hematocrit (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Heart_rate" target="_blank">
<em>HR</em></a> [Heart rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hypokalemia" target="_blank">
<em>K</em></a> [Serum potassium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Lactic_acid" target="_blank">
<em>Lactate</em></a> (mmol/L)</li>
<li><a href="http://en.wikipedia.org/wiki/Magnesium#Biological_role"
target="_blank">
<em>Mg</em></a> [Serum magnesium (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>MAP</em></a> [Invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mechanical_ventilation"
target="_blank">
<em>MechVent</em></a> [Mechanical ventilation respiration (0:false, or 1:true)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_sodium" target="_blank">
<em>Na</em></a> [Serum sodium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>NIDiasABP</em></a> [Non-invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>NIMAP</em></a> [Non-invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>NISysABP</em></a> [Non-invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaCO2</em></a> [partial pressure of arterial CO<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaO2</em></a> [Partial pressure of arterial O<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>pH</em></a> [Arterial pH (0-14)]</li>
<li><a href="http://en.wikipedia.org/wiki/Platelets" target="_blank">
<em>Platelets</em></a> (cells/nL)</li>
<li><a href="http://en.wikipedia.org/wiki/Respiratory_physiology" target="_blank">
<em>RespRate</em></a> [Respiration rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>SaO2</em></a> [O<sub>2</sub> saturation in hemoglobin (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>SysABP</em></a> [Invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Normal_human_body_temperature"
target="_blank">
<em>Temp</em></a> [Temperature (&deg;C)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropI</em></a> [Troponin-I (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropT</em></a> [Troponin-T (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Fluid_balance" target="_blank">
<em>Urine</em></a> [Urine output (mL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Reference_ranges_for_blood_tests#Hematology" target="_blank">
<em>WBC</em></a> [White blood cell count (cells/nL)]</li>
<li>
<em>Weight</em> (kg)</li>


##Accessing the Data
The data frame can be loaded with the following code:

```{r}
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
```

# Task 1 (15 marks)

In this task, you are required to develop a logistic regression model using the `icu_patients_df1` data set which adequately explains or predicts the `in_hospital_death` variable as the outcome using a subset of the available predictor variables. You should fit a series of models, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge. It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct or complete, but do you best). Aim for between five and ten predictor variables (slightly more or fewer is OK). You should assess each model you consider for goodness of fit and other relevant statistics to help you choose between them. For your final model, present a set of diagnostic statistics and/or charts and comment on them. You don't need to do an exhaustive exploratory data analysis of all the variables in the data set, but you should examine those variables that you use in your model. Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find compared to the same model fitted to the imputed data. 


### Hints

1. Select an initial subset of explanatory variables that you will use to predict the risk of in-hospital death. Justify your choice.


```{r}
summary(icu_patients_df1)
```


2. Conduct basic exploratory data analysis on your variables of choice.

3. Fit appropriate univariate logistic regression models.

4. Fit an appropriate series of multivariable logistic regression models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

5. Present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge.

6. For your final model, present a set of diagnostic statistics and/or charts and comment on them.

7. Write a paragraph summarising the most important findings of your final model. Include the most important values from the statistical output, and a simple clinical interpretation.



#####
**Create your response to this task here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text.** 



# Task 2 (15 marks)

In this task, you are required to develop a Cox proportional hazards survival model using the `icu_patients_df1` data set which adequately explains or predicts the length of survival indicated by the `Days` variable, with censoring as indicated by the `Status` variable.  You should fit a series of models, maybe three or four, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge. Aim for between five and ten predictor variables (slightly more or fewer is OK). It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct, but do you best). You should assess each model you consider for goodness of fit and other relevant statistics, and you should assess your final model for violations of assumptions and perform other diagnostics which you think are relevant (and modify the model if indicated, or at comment on the possible impact of what your diagnostics show). Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find.


Import library
```{r}
library(dplyr)
library(survival)
library(survminer)
```


Clean the dataset
```{r}
clean <- function(var1, var2){
  ifelse(is.na(var1), var2, var1)
}

df1_clean <- icu_patients_df1 %>% mutate(DiasABP_diff_clean = clean(DiasABP_diff, NIDiasABP_diff)) %>%
                                  select(-DiasABP_diff, -NIDiasABP_diff) %>%
                                  mutate(DiasABP_max_clean = clean(DiasABP_max, NIDiasABP_max)) %>%
                                  select(-DiasABP_max, -NIDiasABP_max) %>%
                                  mutate(DiasABP_min_clean = clean(DiasABP_min, NIDiasABP_min)) %>%
                                  select(-DiasABP_min, -NIDiasABP_min) %>%
                                  mutate(SysABP_diff_clean = clean(SysABP_diff, NISysABP_diff)) %>%
                                  select(-SysABP_diff, -NISysABP_diff) %>%
                                  mutate(SysABP_max_clean = clean(SysABP_max, NISysABP_max)) %>%
                                  select(-SysABP_max, -NISysABP_max) %>%
                                  mutate(SysABP_min_clean = clean(SysABP_min, NISysABP_min)) %>%
                                  select(-SysABP_min, -NISysABP_min) %>%
                                  mutate(MAP_diff_clean = clean(MAP_diff, NIMAP_diff)) %>%
                                  select(-MAP_diff, -NIMAP_diff) %>%
                                  mutate(MAP_max_clean = clean(MAP_max, NIMAP_max)) %>%
                                  select(-MAP_max, -NIMAP_max) %>%
                                  mutate(MAP_min_clean = clean(MAP_min, NIMAP_min)) %>%
                                  select(-MAP_min, -NIMAP_min) 

```

With regards to blood pressure measurement, invasive measurement is more accuracy than noninvasive mearsurements. Therefore, if invasive blood pressure is recorded, we prefer this record rather than noninvasive record no matter it is recorded or not. However, if  invasive blood pressure is not recorded, we have no choice but choose noninvasive record.


### Hints

1. Select an initial subset of explanatory variables that you will use to predict survival. Justify your choice.

----Descriptor Variables:

Age, Gender, ICUType: demograhpic variables and ICU type will be included in the explanatory variables.

```{r}
coxph(Surv(Days, Status) ~ Age, data = df1_clean)
coxph(Surv(Days, Status) ~ Gender, data = df1_clean)
coxph(Surv(Days, Status) ~ ICUType, data = df1_clean)
```

Results: all of them is statistical significance or close to. 

----Clinical Information Variables:

--Summary indicator

SAPS1: the paper indicates that SAPS correctly classified patients in groups of increasing probability of death. However, it doesn't mention anything in terms of the relationship between SAPS and survival time. From the backgroud knowledge, there is no reason to exclude it from our modeling.

SOFA score: another paper indicates that MOF score correlated to the mortality of critical patients. However, it doesn't mention anything in terms of the relationship between SAPS and survival time. From the backgroud knowledge, there is no reason to exclude it from our modeling.

Conclusion: SAPS1 and SOFA will be part of the initial subset of explanatory variables. However, if the inclusion of SAPS1 and SOFA can not strongly increase the performance of the final model, they will be not in the final model.

--Liver function indicator:

Albumin_max: albumin is a marker for the goodness of liver function and it can be supplemented by injection. Therefore, if the maximum albumin amount were still low after treatment, the survival days of the patients would be shorter.

ALP_min and Bilirubin_min: both of them are markers for not only the badness of liver function but also the severity of hemolysis. If the minimum amout of them were still high after treatment, the survival days of the patients would be shorter.

ALT and AST: both of them are indicators for hepatocellular injury. The reason why it is not selected as initial explanatory variable is that they are easily affected by the medication prescribed by the ICU doctors.

Conclusion: First of all, ALT and AST will be exculsive from the modeling, and the rest of them will be inclusive in the inital analysis. After analysing, one or two of them may remain depending on the results.


```{r}
coxph(Surv(Days, Status) ~ Albumin_max, data = df1_clean)
coxph(Surv(Days, Status) ~ ALP_min, data = df1_clean)
coxph(Surv(Days, Status) ~ Bilirubin_min, data = df1_clean)
coxph(Surv(Days, Status) ~ Albumin_max + ALP_min, data = df1_clean)
coxph(Surv(Days, Status) ~ Albumin_max + Bilirubin_min, data = df1_clean)
```

Results: All of the variables is statistical significance.

```{r}
loglik <- c(summary(coxph(Surv(Days, Status) ~ Albumin_max, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ ALP_min, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ Bilirubin_min, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ Albumin_max + ALP_min, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ Albumin_max + Bilirubin_min, data = df1_clean))$loglik[2])

names(loglik) <- c('Albumin_max','ALP_min','Bilirubin_min','Albumin_max + ALP_min','Albumin_max + Bilirubin_min')

loglik
```

Results: with regards to maximum log-likelihood value, the combination of Albumin_max and Bilirubin_min has better performance. Therefore, Albumin_max and Bilirubin_min will be putted into multivariable model.

--Kidney function indicator:

Creatinine_min ,BUN_min: both of them are marker for the badness of kidney function and it can be decreased by blood purification. Therefore, if the minimum amount of them were still high after treatment, the survival days of the patients would be shorter.

Urine_max: Unrine output is a marker for not only kidney function but also the amount of bleeding. However, it might be not accurate within one-day record. Therefore, it will be exculsive from the initial analysis

Conclusion: Creatinine_min and BUN_min will be inclusive in the inital modeling, and one of them might be exclusive depending on the result from the analysis.

```{r}
coxph(Surv(Days, Status) ~ Creatinine_min, data = df1_clean)
coxph(Surv(Days, Status) ~ BUN_min, data = df1_clean)
coxph(Surv(Days, Status) ~ Creatinine_min + BUN_min, data = df1_clean)
```

Results: All of the variables is statistical significance.

```{r}
loglik <- c(summary(coxph(Surv(Days, Status) ~ Creatinine_min, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ BUN_min, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ Creatinine_min + BUN_min, data = df1_clean))$loglik[2])

names(loglik) <- c('Creatinine_min', 'BUN_min', 'Creatinine_min + BUN_min')

loglik
``` 

Results: with regards to the maximum log-likelihood value, the model including BUN_min has better performance comapred to the model including Creatinine_min, and the performance of the combined model is slightly better than the BUN_min model. Therfore, the BUN_min variable will be putted into multivariable model.

--Central nervous function indicator:

GCS_max: GSC score is a marker for the goodness of central nervous function. If the maximum score were still low after treatment, the survival days of the patients would be shorter.

Conclusion: GCS_max will be inclusive in the initial analysis.

```{r}
coxph(Surv(Days, Status) ~ GCS_max, data = df1_clean)
summary(coxph(Surv(Days, Status) ~ GCS_max, data = df1_clean))$loglik[2]
```

Results: GCS_max will be putted into multivariable model.

--Cardiovascular function indicator:

TropI_max and TropT_max: both of them are markers to assess the severity of myocardial damage, and the concentration of them are proportional to the the severity. If the maximum level of them is pretty high, the risk of death because of heart failure will be increasing, leading to the shorter survival time. Therefore, one of them will be chosed depding on the univariable analysis.

Heart rate: it will be not inclusive in the model analysis because it is easily affected by medication, other treatment and the mood of patient.

MAP_max_clean and MAP_min_clean: both of them are essential to assess the cardiovascular funtion.



```{r}
coxph(Surv(Days, Status) ~ TroponinI_max, data = df1_clean)
coxph(Surv(Days, Status) ~ TroponinT_max, data = df1_clean)
coxph(Surv(Days, Status) ~ MAP_max_clean, data = df1_clean)
coxph(Surv(Days, Status) ~ MAP_min_clean, data = df1_clean)
coxph(Surv(Days, Status) ~ MAP_max_clean + TroponinT_max, data = df1_clean)
coxph(Surv(Days, Status) ~ MAP_min_clean + TroponinT_max, data = df1_clean)
coxph(Surv(Days, Status) ~ MAP_max_clean + MAP_min_clean + TroponinT_max, data = df1_clean)
```

Results: all of the variables except TroponinI_max is statistical significance.

```{r}
loglik <- c(summary(coxph(Surv(Days, Status) ~ TroponinT_max, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ MAP_max_clean, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ MAP_min_clean, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ MAP_max_clean + TroponinT_max, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ MAP_min_clean + TroponinT_max, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ MAP_max_clean + MAP_min_clean + TroponinT_max, 
                          data = df1_clean))$loglik[2])

names(loglik) <- c('TroponinT_max', 'MAP_max_clean', 'MAP_min_clean', 'MAP_max_clean + TroponinT_max', 
                   'MAP_min_clean + TroponinT_max', 'MAP_max_clean + MAP_min_clean + TroponinT_max')

loglik
```

Results: based on the maximum log-likelihood, model consisting of MAP_max_clean, MAP_min_clean and TroponinT_max will be in favour.

--Respiratory function indicator:

RespRate: Respiratory rates are affected by the overall level of PaCO2 and PaO2, therefore, it will no be inclusive in the explanatory variables.

FiO2: almost all of the patients in ICU will get the inspired O2 treatment. Therefore, this variable is not trustful.

SaO2: it is affected by PaO2 in most of the time.

PaO2_max: PaO2 is a marker for the goodness of the respiratory function. If the maximum amount of PaO2 is still low after treatment, the survival time of a patient will be shorter.

PaCO2_diff: PaCO2 is a sensitive marker for assessing gas exchange inside the lungs. If the extreme value of PaCO2 is far away from the mean value, it means higher probability of a patient lack of breathing.

```{r}
coxph(Surv(Days, Status) ~ PaO2_max, data = df1_clean)
coxph(Surv(Days, Status) ~ PaCO2_diff, data = df1_clean)
coxph(Surv(Days, Status) ~ PaO2_max + PaCO2_diff, data = df1_clean)
```

Results:  All of the variables is statistical significance.

```{r}
loglik <- c(summary(coxph(Surv(Days, Status) ~ PaO2_max, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ PaCO2_diff, data = df1_clean))$loglik[2],
            summary(coxph(Surv(Days, Status) ~ PaO2_max + PaCO2_diff, data = df1_clean))$loglik[2])

names(loglik) <- c('PaO2_max', 'PaCO2_diff', 'PaO2_max + PaCO2_diff')

loglik
```

Results: with regards to the maximum log-likelihood value, model combined of PaO2_max and PaCO2_diff has the best performance. Therefore, PaO2_max and PaCO2_diff will be putted into multivariable model.

--Coagulation function indicator:

Platelets_max: platelets ia a marker for the goodness of coagulation function and it can be supplemented by injection. Therefore, if the maximum platelets amount were still low after treatment, the survival days of the patients would be shorter.

```{r}
coxph(Surv(Days, Status) ~ Platelets_max, data = df1_clean)
summary(coxph(Surv(Days, Status) ~ Platelets_max, data = df1_clean))$loglik[2]
```

Results: although Platelets_max is not statistical significance, it will be still part of the multivariable because it is the only variable provided to assess coagulation function.

--Infection indicator

WBC_min: it is a indicator for the severity of infection. If the minimum amount of WBC is still high after using antibiotic, then the survival time will be shorter.

body temperature: it is less trustful compared to WBC because it might be affected by the enviornmental tempeture, BMI, stress reaction and adverse effect of medications.

--Blood biochemistry results

Blood biochemistry results in one-day record will be less trustful. Morever, the blood biochemistry results are some kind of comprehensive results from the liver function, kidney funciton and respirtory function. Therefore, they will be not in the explanatory variables.


```{r}
summary(coxph(Surv(Days, Status) ~ Age + Gender + ICUType + Albumin_max + Bilirubin_min + BUN_min + 
                GCS_max + MAP_max_clean + MAP_min_clean + TroponinT_max + PaO2_max + 
                PaCO2_diff + Platelets_max + WBC_min + SAPS1 + SOFA, data = df1_clean))
```











2. Conduct basic exploratory data analysis on your variables of choice.

```{r}
df1_clean %>% select(Age, Gender, ICUType, Albumin_max, Bilirubin_min, BUN_min,
                GCS_max, MAP_max_clean, MAP_min_clean, TroponinT_max, PaO2_max,
                PaCO2_diff, Platelets_max, WBC_min, SAPS1, SOFA) %>% summary
```

Obviously the range of SOFA variable is out of range cause its allowable entry is ranging from 0 to 24. Therefore, some cleaning work is required.

```{r}
df1_clean[df1_clean$SOFA == -1,]$SOFA = NA
```


3. Fit appropriate univariate Cox proportional hazards models.

4. Fit an appropriate series of multivariable Cox proportional hazards models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

5. Present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge.

6. For your final model, present a set of diagnostic statistics and/or charts and comment on them. 

7. Write a very brief paragraph summarising the most important findings of your final model. Include the most important values from the statistical output, and a simple clinical interpretation. 

####
**Create your response to this task here, as a mixture of embedded (`knitr`) R code and any resulting outputs, and explanatory or commentary text.** 



## Save, knit and submit

**Reminder**: don't forget to save this file, to knit it to check that everything works, and then submit via the drop box in Openlearning.

## Submit your assignment

When you have finished, and are satisfied with your assignment solutions, and this file knits without errors and the output looks the way you want, then you should submit via the drop box in Openlearning.

### Problems?

If you encounter problems with any part of the process described above, please contact the course instructor via OpenLearning as soon as possible so that the issues can be resolved in good time, and well before the assignment is due.


### Additional Information

Each task attracts the indicated number of marks (out of a total of 30 marks for the assignment). The instructions are deliberately open-ended and less prescriptive than the individual assignments to allow you some latitude in what you do and how you go about the task. However, to complete the tasks and gain full marks, you only need to replicate or repeat the steps covered in the course - if you do most or all of the things described in the revalant chapters of the HDAT9600 course, full marks wil be awarded. 

Note also that with respect to the model fitting, there are no **right** or **wrong** answers when it comes to variable selection and other aspects of model specification. Deep understanding of the underlying medical concepts which govern patient treatment and outcomes in ICUs is not required or assumed, although you should try to gain some understanding of each variable using the links provided. You will not be marked down if your medical justifications are not exactly correct or complete, but do you best, and don't hesitate to seek help from the course instructor.
