---
title: "HDAT9600 Final Assignment"
author: "insert your team name here"
date: "insert date of completion here"
output:
  pdf_document: default
  html_document: default
subtitle: 'Submission deadline is 5pm Friday 23 2019 '
---

```{r setup, include=FALSE}
# leave this code here, but feel free to adjust the options or add some more
# see the knitr documentation for details
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=12)
```

## Instructions

This file (hdat9600_final_assignment.Rmd) is the R Markdown document in which you need to complete your HDAT9600 final assignment. This assignment is assessed and will count for 30% of the total course marks. The assignment comprises two tasks worth 15 marks each. The first task will focus on logistic regression, and the second task will focus on survival analysis. There is no word limit, but a report of about 10 pages in length when printed (except that it will not be printed) is appropriate.

Don't hesitate to ask the course convenor for help via OpenLearning. The course instructor are happy to point you in the right direction and to make suggestions, but they won't, of course, complete your assignments for you!


## Data for this assignment

The data used for this assignment consist of records from Intensive Care Unit (ICU) hospital stays in the USA. All patients were adults who were admitted for a wide variety of reasons. ICU stays of less than 48 hours have been excluded. 

The source data for the assignment are data made freely available for the 2012 MIT PhysioNet/Computing for Cardiology Challenge. Details are provided [here](https://physionet.org/challenge/2012/). Training Set A data have been used. THe original data has been modified and assembled to suit the purpose of this assignment. Full details of the preparatory work can be found in the hdat9600-final-assignment-data-preparation file.

The dataframe consists of 120 variables, which are defined as follows:

####Patient Descriptor Variables
<li> <em>RecordID:</em>   a unique integer for each ICU stay</li>
<li> <em>Age:</em>                 years</li>
<li> <em>Gender:</em>              male/female</li>
<li> <em>Height:</em>              cm</li>
<li> <em>ICUType:</em>             Coronary Care Unit; Cardiac Surgery Recovery Unit; Medical ICU; Surgical ICU</li>
<li> <em>Length_of_stay:</em>      The number of days between the patient’s admission to the ICU and the end of hospitalisation</li>
<li> <em>Survival:</em>            The number of days between ICU admission and death for patients who died</li>


####Outcome Variables
<li> <em>in_hospital_death:</em>   0:survivor/1:died in-hospital **this is the outcome variable for Task 1: Logistic Regression**</li>
<li> <em>Status:</em>              True/False **this is the censoring variable for Task 2: Survival Analysis**</li>
<li> <em>Days:</em>                Length of survival (in days) **this is the survival time variable for Task 2: Survival Analysis**</li>


####Clinical Variables
<em>Use the hyperlinks below to find out more about the clinical meaning of each variable.</em>
The first two clinical variables are summary scores that are used to assess patient condition and risk.

<li> <em>SAPS-I score</em> [Simplified Acute Physiological Score (<a href="http://www.ncbi.nlm.nih.gov/pubmed/6499483"
target="_blank"><em>Le Gall et al., 1984</em></a>)]</li>
<li> <em>SOFA score</em> [Sequential Organ Failure Assessment (<a href="http://www.ncbi.nlm.nih.gov/pubmed/11594901"
target="_blank"><em>Ferreira et al., 2001</em></a>)]</li>

###

The following 36 clinical measures were assessed at multiple timepoints during each patient's ICU stay. For each of the 36 clinical measures, you are given 3 summary variables: a) The minimum value during the first 24 hours in ICU (_min), b) The maximum value during the first 24 hours in ICU (_max), and c) The difference between the mean and the most extreme values during the first 24 hours in ICU (_diff). For example, for the clinical measure Cholesterol, these three variables are labelled 'Cholesterol_min', 'Cholesterol_max', and 'Cholesterol_diff'.

<li><a href="http://en.wikipedia.org/wiki/Human_serum_albumin" target="_blank">
<em>Albumin</em></a> (g/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Alkaline_phosphatase" target="_blank">
<em>ALP</em></a> [Alkaline phosphatase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Alanine_transaminase" target="_blank">
<em>ALT</em></a> [Alanine transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Aspartate_transaminase" target="_blank">
<em>AST</em></a> [Aspartate transaminase (IU/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bilirubin" target="_blank">
<em>Bilirubin</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/BUN" target="_blank">
<em>BUN</em></a> [Blood urea nitrogen (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Cholesterol" target="_blank">
<em>Cholesterol</em></a> (mg/dL)</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_creatinine#Plasma_creatinine"
target="_blank">
<em>Creatinine</em></a> [Serum creatinine (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>DiasABP</em></a> [Invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/FIO2" target="_blank">
<em>FiO2</em></a> [Fractional inspired O<sub>2</sub> (0-1)]</li>
<li><a href="http://en.wikipedia.org/wiki/Glasgow_coma_score" target="_blank">
<em>GCS</em></a> [Glasgow Coma Score (3-15)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_glucose" target="_blank">
<em>Glucose</em></a> [Serum glucose (mg/dL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Bicarbonate#Diagnostics"
target="_blank">
<em>HCO3</em></a> [Serum bicarbonate (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hematocrit" target="_blank">
<em>HCT</em></a> [Hematocrit (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Heart_rate" target="_blank">
<em>HR</em></a> [Heart rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Hypokalemia" target="_blank">
<em>K</em></a> [Serum potassium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Lactic_acid" target="_blank">
<em>Lactate</em></a> (mmol/L)</li>
<li><a href="http://en.wikipedia.org/wiki/Magnesium#Biological_role"
target="_blank">
<em>Mg</em></a> [Serum magnesium (mmol/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>MAP</em></a> [Invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mechanical_ventilation"
target="_blank">
<em>MechVent</em></a> [Mechanical ventilation respiration (0:false, or 1:true)]</li>
<li><a href="http://en.wikipedia.org/wiki/Serum_sodium" target="_blank">
<em>Na</em></a> [Serum sodium (mEq/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Diastolic_blood_pressure"
target="_blank">
<em>NIDiasABP</em></a> [Non-invasive diastolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Mean_arterial_pressure">
<em>NIMAP</em></a> [Non-invasive mean arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>NISysABP</em></a> [Non-invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaCO2</em></a> [partial pressure of arterial CO<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>PaO2</em></a> [Partial pressure of arterial O<sub>2</sub> (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>pH</em></a> [Arterial pH (0-14)]</li>
<li><a href="http://en.wikipedia.org/wiki/Platelets" target="_blank">
<em>Platelets</em></a> (cells/nL)</li>
<li><a href="http://en.wikipedia.org/wiki/Respiratory_physiology" target="_blank">
<em>RespRate</em></a> [Respiration rate (bpm)]</li>
<li><a href="http://en.wikipedia.org/wiki/Arterial_blood_gas" target="_blank">
<em>SaO2</em></a> [O<sub>2</sub> saturation in hemoglobin (%)]</li>
<li><a href="http://en.wikipedia.org/wiki/Systolic_blood_pressure"
target="_blank">
<em>SysABP</em></a> [Invasive systolic arterial blood pressure (mmHg)]</li>
<li><a href="http://en.wikipedia.org/wiki/Normal_human_body_temperature"
target="_blank">
<em>Temp</em></a> [Temperature (&deg;C)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropI</em></a> [Troponin-I (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Troponin" target="_blank">
<em>TropT</em></a> [Troponin-T (&mu;g/L)]</li>
<li><a href="http://en.wikipedia.org/wiki/Fluid_balance" target="_blank">
<em>Urine</em></a> [Urine output (mL)]</li>
<li><a href="http://en.wikipedia.org/wiki/Reference_ranges_for_blood_tests#Hematology" target="_blank">
<em>WBC</em></a> [White blood cell count (cells/nL)]</li>
<li>
<em>Weight</em> (kg)</li>

##Load Libraries

```{r warning=FALSE, include=FALSE}
#Set up Libraries
list.of.packages <- c("ggplot2", "Rcpp", "DataExplorer", "skimr", "knitr", "DataExplorer")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)


library(tidyverse)
library(magrittr)
library(DataExplorer)
library(skimr)
library(knitr)
library(DataExplorer)

```

## Accessing the Data

```{r echo=TRUE}
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
```

# Task 1 (15 marks)

In this task, you are required to develop a logistic regression model using the `icu_patients_df1` data set which adequately explains or predicts the `in_hospital_death` variable as the outcome using a subset of the available predictor variables. You should fit a series of models, evaluating each one, before you present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge. It is perfectly acceptable to include predictor variables in your final model which are not statistically significant, as long as you justify their inclusion on medical or physiological grounds (you will not be marked down if your medical justification is not exactly correct or complete, but do you best). Aim for between five and ten predictor variables (slightly more or fewer is OK). You should assess each model you consider for goodness of fit and other relevant statistics to help you choose between them. For your final model, present a set of diagnostic statistics and/or charts and comment on them. You don't need to do an exhaustive exploratory data analysis of all the variables in the data set, but you should examine those variables that you use in your model. Finally, re-fit your final model to the unimputed data frame (`icu_patients_df0.rds`) and comment on any differences you find compared to the same model fitted to the imputed data. 

## GLM Task

1. Conduct basic exploratory data analysis on your variables of choice.

_First objectives are to assess the properties of our data, assess the data quality, and lastly determine which variables are correlated or clinically relevant_

Key take aways from the data prep explaination:

* Patients with a total ICU stay less than 48 hours have been excluded. Of these patients we are looking at only the first 24 hours of the patient's stay in ICU
* Patients with DNR (do not resuscitate) or CMO (comfort measures only) directives were not excluded.
* Up to 42 variables were recorded at least once during the first 48 hours after admission to the ICU. Not all variables are available in all cases, however. Six of these variables are general descriptors (collected on admission), and the remainder are time series, for which multiple observations may be available.
..* Observations with multiple measures are summarised to give maximum, minimum values etc. for the first 24 hours in ICU.
..* Observations captured at admission are:
....1 RecordID (a unique integer for each ICU stay)



###Initial look at our data


```{r results='asis'}
options(max.print=999999)
skim(icu_patients_df1) %>% skimr::kable()
```

_Most variables have been cleaned using the preliminary data-prep associated with this assignment. However, we can see that most of the continuous variables contain a few extreme maximum values and the majority of the data is not missing. A few specific variables such as DiasABP, Height, NI-pathology results, and SysABP measures are worth investigating._


_We can also drop columns such as survival and status which are not related to the GLM analysis._

```{r}
icu_patients_df1 <- subset(icu_patients_df1, select = -c(Status, Survival, Days, Length_of_stay))
```


```{r}
missing_rates<- profile_missing(icu_patients_df1) %>% arrange(desc(num_missing))

missing_rates[1:25,]
```


###Check Our discrete variables


```{r}
plot_bar(icu_patients_df1)
```


###Check continuous variables


```{r}
plot_histogram(icu_patients_df1)
```

2. Select an initial subset of explanatory variables that you will use to predict the risk of in-hospital death. Justify your choice.



###Look at discrete variable correlations





```{r}
plot_correlation(na.omit(icu_patients_df1), maxcat = 5L, type = "d")
```

_Most ICU admissions are negatively correlated with other ICU admissions. This is intuative as it is rare to require multiple different ICU admissions except in complex patients. Cardiac Surgery Recovery ICU and Surgical ICU are the most negatively correlated at -0.46._

```{r}
plot_correlation(na.omit(icu_patients_df1), maxcat = 5L, type = "c")
```

A few key relationships stick out between variables looking at this graph:

* The 'Creatinine' variables (min, dif, max) are positively correlated with corresponding 'BUN' Blood Urea Nitrogen variables.
* Systolic ABP values are positively correlated with corresponding Diasystolic ABP values with corresponding MAP values (especially the minimum values).



From this graph we can make a preliminary assessement about which predictor variables will be suitable for logistic regression analysis. We can select for predictor variables which are not highly correlated with each other, but are correlated with our 'in_hospital_death' outcome variable.

1. Albumin min -

*no strong correlations between other predictor variables, other than other ablumin variables eg. Albumin_max - which we will not include.

2. Bilirubin max - Bilirubin_max 

* Clinical indicator of impaired renal function and Haemolysis

3. Serum creatinine (mg/dL) - Creatinine_max

* High levels (above 250) are a clincial indicator of renal failure, which is a very serious complication somtimes requiring dialysis.
* strongly Correlated with BUN but also weakly correlated with in_hospital_death.

4. Serum bicarbonate (mmol/L) - HCO3_diff 

* Indicative of blood acidosis, higher Co

5. SAPS-I score - SAPS1

* This scoring system was created using 14 biological/clincial indicators Specifically for ICU paitents facilitate multicentre studies and outcome comparisons.
* Positively correlated with in_hospital_death.

6. Blood urea nitrogen (mg/dL) - BUN min


7. Glasgow Comas Score - GCS_max

* The Glasgow Coma Scale provides a score in the range 3-15; patients with scores of 3-8 are usually said to be in a coma.
* Strongly Negatively correlated with in_hospital_death

8. Arterial pH (0-14) - ph_min

* Can be an indicator of Respiratory acidosis

9. Serum glucose (mg/dL) - Glucose_max

* Clinical indicator of diabetes
* Positive correlation with in_hospital_death.

10. Lactate (mmol/L) - Lactate

* Positive correlation with in_hospital_death.

11. Temperature (°C) - temp_diff

* Large differences in temperature during a patient's stay in ICU are indicative of many underlying issues such as infection.

12. O2 saturation in hemoglobin (%) - SaO2_max

* Haemoglobin oxygen saturation
* Negatively correlated with in_hospital_death.

13. Height (cm) - Height

* Standard to include within medical modelling. However, due to the numerous missing values, it will not be included.

14. Weight (kg) - Weight

* Greatest difference in weight throughout the stay
* Standard to include within medical modelling.


###Scatterplot 

```{r}
arr_delay_df2 <- icu_patients_df1[, c("in_hospital_death", "ICUType", "SAPS1", "SOFA", "Albumin_min", "Bilirubin_max", "Creatinine_max", "HCO3_diff", "SAPS1", "BUN_min", "GCS_max", "pH_min", "Glucose_max", "SaO2_max", "Weight_diff", "Urine_diff")]

plot_scatterplot(arr_delay_df2, by = "in_hospital_death", sampled_rows = 1000L)
```

_investigation of relationships between "in_hospital_death" and chosen predictor variables._


3. Fit appropriate univariate logistic regression models.

###Univariate

```{r echo=TRUE}
#put column names into a variable 
categorical_vars <- c("ICUType", "SAPS1", "SOFA", "Albumin_min", "Bilirubin_max", "Creatinine_max", "HCO3_diff", "SAPS1", "BUN_min", "GCS_max", "pH_min", "Glucose_max", "SaO2_max", "Weight_diff", "Urine_diff")

univariate_mods <- c() #initialise an empty list of vectors for the loop

for(i in categorical_vars){
  modelformula <- paste0("in_hospital_death ~ ", i) #use for loop iterator to create a formula for each variable
  univariate_mods[[i]] <- glm(as.formula(modelformula), data = icu_patients_df1, family = "binomial")
}

for(i in univariate_mods){
  print(summary(i))
}
```

_SaO2, Weight, and Height Variables were not sigificant. Retain Weight and Height as these are standard to include in medical modelling._

4. Fit an appropriate series of multivariable logistic regression models, justifying your approach. Assess each model you consider for goodness of fit and other relevant statistics.

###Multivariate

```{r}
#put column names into a variable 
categorical_vars <- paste(c("ICUType", "SAPS1", "SOFA", "Albumin_min", "Bilirubin_max", "Creatinine_max", "HCO3_diff", "SAPS1", "BUN_min", "GCS_max", "pH_min", "Glucose_max", "SaO2_max", "Weight_diff", "Urine_diff"), collapse = " + ")

modelformula <- paste("in_hospital_death ~ ", categorical_vars)

multivariate_mod <- glm(modelformula, data = icu_patients_df1, family = "binomial")

summary(multivariate_mod)
```

5. Present your final model. Your final model should **not** include all the predictor variables, just a small subset of them, which you have selected based on statistical significance and/or background knowledge.

```{r}

```




6. For your final model, present a set of diagnostic statistics and/or charts and comment on them.

7. Write a paragraph summarising the most important findings of your final model. Include the most important values from the statistical output, and a simple clinical interpretation.




#NOTES

- we may want to revise variable selection based upon some predictor variable correlations eg. BUN and creatinine.




## Save, knit and submit

**Reminder**: don't forget to save this file, to knit it to check that everything works, and then submit via the drop box in Openlearning.

## Submit your assignment

When you have finished, and are satisfied with your assignment solutions, and this file knits without errors and the output looks the way you want, then you should submit via the drop box in Openlearning.

### Problems?

If you encounter problems with any part of the process described above, please contact the course instructor via OpenLearning as soon as possible so that the issues can be resolved in good time, and well before the assignment is due.


### Additional Information

Each task attracts the indicated number of marks (out of a total of 30 marks for the assignment). The instructions are deliberately open-ended and less prescriptive than the individual assignments to allow you some latitude in what you do and how you go about the task. However, to complete the tasks and gain full marks, you only need to replicate or repeat the steps covered in the course - if you do most or all of the things described in the revalant chapters of the HDAT9600 course, full marks wil be awarded. 

Note also that with respect to the model fitting, there are no **right** or **wrong** answers when it comes to variable selection and other aspects of model specification. Deep understanding of the underlying medical concepts which govern patient treatment and outcomes in ICUs is not required or assumed, although you should try to gain some understanding of each variable using the links provided. You will not be marked down if your medical justifications are not exactly correct or complete, but do you best, and don't hesitate to seek help from the course instructor.
