---
title: "Rory Final Assessment HDAT9600 EDA"
author: "Rory Denham"
date: "06/08/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Set up Libraries
list.of.packages <- c("ggplot2", "Rcpp", "DataExplorer")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)


library(tidyverse)
library(magrittr)
library(DataExplorer)
```

```{r}
# read the outcomes CSV file
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
```

```{r}
str(icu_patients_df0)
```

```{r}
introduce(icu_patients_df1)
plot_intro(icu_patients_df1)
```

_Here we can see that most of our variables are Continuous and only 2.5 percent of our variables are Factored._

_Additionally we can see that only 9.2% of the observations within our dataset are complete (ie. each column for a given row has a value). However, there are also only 4.5% 

```{r}
plot_missing(icu_patients_df1)
```

Survival and Height are contain a very large amount of missing variables within our dataset. We should investigate these variables to understand their structure better. It may be the case that NA is equivalent to suriving.

```{r}
plot_bar(icu_patients_df1)
```

```{r}
plot_bar(icu_patients_df1)
```

```{r}
plot_histogram(icu_patients_df1)
```

```{r}

```

